---
title: "Spread of a pathogen in a discrete structured population"
author: "Sebastian Lequime"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{single-host-discrete}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Aside from the simple simulation set up, explored in [another tutorial](none.html), where hosts are "not structured", `nosoi` can take into account a population structured either in discrete states or in a continuous space. We will here focus on a discrete state structuration (for the continuous structuration, see [this tutorial](continuous.html)).

The discrete structuration is intended to allow the simulation to take place into a geographical context between discrete states such as countries, regions, islands, cities. However, other kind of structure could be taken into account (high risk/low risk, etc). This also allows parameter values to change according to the host's current location among the available states. This tutorial will focus on setting up a `nosoi` simulation for a pathogen whose host population is structured between different locations.

# Structure of the population

We will consider here a population of hosts that are in three different locations called "A", "B" and "C". Hosts can move (if they undergo a movement event) between these locations with a certain probability you can set or derive from other data. Here, the transition matrix, called here after `structure.matrix` is:
```{r setupMatrix2, echo=FALSE}
matrix(c(0,0.2,0.4,0.5,0,0.6,0.5,0.8,0),nrow = 3, ncol = 3,dimnames=list(c("A","B","C"),c("A","B","C")))
```

Which can graphically be represented:

```{r setupMatrix, echo=FALSE, message=FALSE,warning=FALSE}
library(ggplot2)
library(viridis)
library(igraph)
library(ggnetwork)

transition.matrix = matrix(c(0,0.2,0.4,0.5,0,0.6,0.5,0.8,0),nrow = 3, ncol = 3,dimnames=list(c("A","B","C"),c("A","B","C")))

melted.transition.matrix <- reshape2::melt(transition.matrix, varnames = c("from","to"),value.name="prob", as.is = TRUE) #melting the matrix go get from -> to in one line with probability

melted.transition.matrix = subset(melted.transition.matrix, prob!=0)

graph.Matrix <- graph.data.frame(melted.transition.matrix,directed=T)

graph.matrix.network = ggnetwork(graph.Matrix, layout = "circle",arrow.gap=0.18) #using ggnetwork to provide the layout

#plotting the network
ggplot(graph.matrix.network, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_edges(color = "grey70",arrow = arrow(length = unit(1, "lines"), type = "closed"),curvature = 0.2) + geom_nodes(aes(color =vertex.names) , size = 30) +
  geom_nodetext(aes(label = vertex.names),color="white", fontface = "bold",size=15) + scale_color_viridis(guide=FALSE,discrete=TRUE) +
  theme_blank() + ylim(-0.5,1.2) + xlim(-0.5,1.2) +
  geom_text(data=graph.matrix.network,aes(x=(x+xend)/2,y=(y+yend)/2,label = prob,color=vertex.names), size = 6)
```

For `structure.matrix` to be adequate, a few rules have to be followed:

- `structure.matrix` should be of class `matrix`;
- `structure.matrix` should have the same number of rows and columns, and they should have the same names;
- `structure.matrix` rows represents departure state and columns arrival states, the value is the probability from going from departure state to arrival state if movement is undergone, and should thus sum up to 1.

# Setting up the simulation

The wrapper function `nosoiSim` takes all the arguments that will be passed down to the simulator, in the case of this tutorial `singleDiscrete` (for "single host, discrete structure").
We thus start by providing the options `type="single"` and `popStructure="discrete"` to set up the analysis:

```{r setupA, eval = FALSE}
SimulationA <- nosoiSim(type="single", popStructure="discrete", ...)
```

This simulation type takes several arguments or options to be able to run, namely:

- `length.sim` 
- `max.infected`
- `init.individuals`
- `init.structure`
- `structure.matrix`
- `pExit` with `param.pExit`, `timeDep.pExit` and `diff.pExit` 
- `pMove` with `param.pMove`, `timeDep.pMove` and `diff.pMove`
- `nContact` with `param.nContact`, `timeDep.nContact` and `diff.nContact`
- `pTrans` with `param.pTrans`, `timeDep.pTrans` and `diff.pTrans`
- `prefix.host.A`
- `print.progress`
- `print.step`

All the `param` elements provide individual-level parameters to be taken into account, while the `timeDep` elements inform the simulator if the "absolute" simulation time should be taken into account. The `diff` elements inform the simulator if there is a differential probability according to the state the host is currently in.

## General parameters

`length.sim`, `max.infected` and `init.individuals` are general parameters that define the simulation.
`length.sim` is the maximum number of time units (e.g. days, months, years, or another time unit of choice) during which the simulation will be run, whereas `max.infected` is the maximum number of individuals that can be infected during the simulation.
`init.individuals` and `init.structure` are the “seeding parameters”. `init.individuals` defines the number of individuals (starting at 1, should be an integer) that will start a transmission chain and `init.structure` their original location in the structured population (has to be the same original location for all starting individuals). Keep in mind that you will have as many transmission chains as initial individuals, which is equivalent as launching two independant nosoi simulations. The location provided in `init.structure` should of course be present in the `structure.matrix`.

Here, we will run a simulation starting with 1 individual, for a maximum of 1.000 infected individuals and a maximum time of 300 days.

```{r setupB, eval = FALSE}
SimulationA <- nosoiSim(type="single", popStructure="none",
                        length.sim=300, max.infected=1000, init.individuals=1, ...)
```

## Core functions

The core functions `pExit`, `pMove`, `nContact` and `pTrans` each follow the [same principles to be set up](nosoi.html#setting-up-the-core-functions). 
To accommodate several scenarios, they can be constant, time-dependent (relative time since infection for each individual or "absolute"" time of the simulation), state-dependent, or even individually parameterized, to include some stochasticity at the individual-host level.
In every case, the function provided, like all other core function in `nosoi`, has to be expressed as function of time `t`, even if time is not used to compute the probability.
In case the function uses individual-based parameters, you have to specify a list of functions (called `param.pExit`, `param.pMove`, `param.nContact` or `param.pTrans`) ( [see Get started](nosoi.html#parameters) ). If no individual-based parameter is used, then they are set to `NA`.

> Keep in mind that `pExit`, `pMove`, and `pTrans` have to give back a probability (i.e. a number between 0 and 1) while `nContact` should give back a positive natural number (positive integer).

You can combine the influence of several parameters, such as time since infection, "absolute" time of the simulation, location (in the discrete states), and individual-based parameters within the same function.

> In any case, time since infection, "absolute" time, and current location should ALWAYS be designated by `t`, `prestime` and `current.in` respectively, and be in that order and before of the individual-based parameters.

### `pExit`, `param.pExit`, `timeDep.pExit` and `diff.pExit`

`pExit` is the first required fundamental parameter and provides a daily probability for a host to leave the simulation (either cured, died, etc.).
`param.pExit` is the list of function need to individually paramterize `pExit` ( [see Get started](nosoi.html#parameters) ). The name of each function in the list has to have the same name as the parameter it is sampling for `pExit`. `timeDep.pExit` allows for `pExit` to be dependant of the "absolute" time of the simulation, to account, for example, for seasonality or other external time related covariates. By default, `timeDep.pExit` is set to `FALSE`.
`diff.pExit` allows `pExit` to be different according to the current discrete state of the host. By default, `diff.pExit` is set to `FALSE`.

### `pMove`, `param.pMove`, `timeDep.pMove` and `diff.pMove`

`pMove` is the probability (per unit of time) for a host to move to a different state (for example, leaving state "A"). It should not be confused with the probabilities extracted from the `structure.matrix`, which represent the probability to go to a specific location once a movement is ongoing (for example, going to "B" or "C" while coming from "A").

`param.pMove` is the list of function need to individually paramterize `pMove` ( [see Get started](nosoi.html#parameters) ). The name of each function in the list has to have the same name as the parameter it is sampling for `pMove`. `timeDep.pMove` allows for `pMove` to be dependant of the "absolute" time of the simulation, to account, for example, for seasonality or other external time related covariates. By default, `timeDep.pMove` is set to `FALSE`.
`diff.pMove` allows `pMove` to be different according to the current discrete state of the host. By default, `diff.pMove` is set to `FALSE`.

### `nContact`, `param.nContact`, `timeDep.nContact` and `diff.nContact`

``nContact` represents the number (a positive integer) of potentially infectious contacts an infected hosts can encounter per unit of time. At each time point, a number of contacts will be determined for each active host in the simulation.
The number of contacts (i.e. the output of your function) has to be an integer and can be set to zero.
`param.nContact` is the list of function need to individually paramterize `nContact` ( [see Get started](nosoi.html#parameters) ). The name of each function in the list has to have the same name as the parameter it is sampling for `nContact`.
`timeDep.nContact` allows for `nContact` to be dependant of the "absolute" time of the simulation, to account, for example, for seasonality or other external time related covariates. By default, `timeDep.nContact` is set to `FALSE`. 
`diff.nContact` allows `nContact` to be different according to the current discrete state of the host. By default, `diff.nContact` is set to `FALSE`.

### `pTrans`, `param.pTrans`, `timeDep.pTrans` and `diff.pTrans`

`pTrans` is the heart of the transmission process and represents the probability of transmission over time (when a contact occurs).
`param.pTrans` is the list of function need to individually paramterize `pTrans` ( [see Get started](nosoi.html#parameters) ). The name of each function in the list has to have the same name as the parameter it is sampling for `pTrans`. 
`timeDep.pTrans` allows for `pTrans` to be dependant of the "absolute" time of the simulation, to account, for example, for seasonality or other external time related covariates. By default, `timeDep.pTrans` is set to `FALSE`. 
`diff.pTrans` allows `pTrans` to be different according to the current discrete state of the host. By default, `diff.pTrans` is set to `FALSE`.

## Miscellaneous

`prefix.host` allows you to define the first character(s) for the hosts' unique ID.
It will be followed by a hyphen and a unique number.
By default, `prefix.host` is "H" for "Host".

`print.progress` allows you to get some information printed on the screen about the simulation while running. It will print something every `print.step`. By default, `print.progress` is activated, with a `print.step = 10` (that you can change), but you may want to deactivate it by providing `print.progress == FALSE`.

## Dual host

In the case of a dual host simulation, several parameter of the `nosoiSim` will have to be specified for each host type, called `A` and `B`. The wrapper function `nosoiSim` will then take all the arguments that will be passed down to the simulator, in the case of this tutorial `dualDiscrete` (for "dual host, discrete structure").
We thus start by providing the options `type="dual"` and `popStructure="discrete"` to set up the analysis:

```{r setupA-dual, eval = FALSE}
SimulationA <- nosoiSim(type="dual", popStructure="discrete", ...)
```

This function takes several arguments or options to be able to run, namely:

- `length.sim` 
- `max.infected.A`
- `max.infected.B`
- `init.individuals.A`
- `init.individuals.B`
- `init.structure.A`
- `init.structure.B`
- `structure.matrix.A`
- `structure.matrix.B`
- `pExit.A` with `param.pExit.A`, `timeDep.pExit.A` and `diff.pExit.A` 
- `pMove.A` with `param.pMove.A`, `timeDep.pMove.A` and `diff.pMove.A`
- `nContact.A` with `param.nContact.A`, `timeDep.nContact.A` and `diff.nContact.A`
- `pTrans.A` with `param.pTrans.A`, `timeDep.pTrans.A` and `diff.pTrans.A`
- `prefix.host.A`
- `pExit.B` with `param.pExit.B`, `timeDep.pExit.B` and `diff.pExit.B` 
- `pMove.B` with `param.pMove.B`, `timeDep.pMove.B` and `diff.pMove.B`
- `nContact.B` with `param.nContact.B`, `timeDep.nContact.B` and `diff.nContact.B`
- `pTrans.B` with `param.pTrans.B`, `timeDep.pTrans.B` and `diff.pTrans.B`
- `prefix.host.B`
- `print.progress`
- `print.step`

As you can see, host-type dependent parameters are now designated by the suffix `.A` or `.B`.

Both `max.infected.A` and `max.infected.B` have to be provided to set an upper limit on the simulation size. To initiate the simulation, you have to provide at least on starting host, either `A` or `B` in `init.individuals.A` or `init.individuals.B` respectively, as well as starting position in `init.individuals.A` or `init.individuals.B`, respectively. If you want to start the simulation with one host only, then `init.individuals` of the other can be set to 0 and `init.structure` to `NA`.

A major difference here is that hosts may, or not, share the same `structure.matrix`. However, since they exist in the same "world", they should share the same states names. It is also possible to have a host that does not move. In such case, `pMove` can be set to `NA`. In such "non-moving" case, a transition matrix, with all states names should still be provided.

# Running `nosoi`

## Single host

We will present here a very simple simulation for a single host pathogen.

### pExit

For `pExit`, we will chose a probability that depends on the location where the host currently is. Each location (state) has to be present, such as this example:

```{r pExit1, eval = FALSE}
p_Exit_fct  <- function(t,current.in){
  if(current.in=="A"){return(0.02)}
  if(current.in=="B"){return(0.05)}
  if(current.in=="C"){return(0.1)}}
```

This function means that if the host is in state "A", it has 2% chance to exit, 5% if in state "B" and 10% if in state "C". Remember that `pExit`, like the other core function has to be function of `t`, even if `t` is not used. Since `pExit` is dependent on the location, `diff.pExit=TRUE`. However, there is no use of the "absolute" time of the simulation nor individual-based parameters, so `param.pExit = NA`, and `timeDep.pExit=FALSE`.

### pMove

We will chose a constant value for `pMove`, namely 0.1, i.e. an infected host has 10% chance to change its state (here a location) for each unit of time.

```{r pMove1, eval = FALSE}
p_Move_fct  <- function(t){return(0.1)}
```

Remember that `pMove`, like the other core function has to be function of `t`, even if `t` is not used. Since `pMove` is not dependent on the location, `diff.pMove=FALSE`. Similarly, there is no use of the "absolute" time of the simulation nor individual-based parameters, so `param.pMove = NA`, and `timeDep.pMove=FALSE`.

### nContact

For `nContact`, we will chose a constant function that will draw a value in a normal distribution of *mean* 0.5 and *sd* 1, round it, and take its absolute value.

```{r nContact1, eval = FALSE}
n_contact_fct = function(t){abs(round(rnorm(1, 0.5, 1), 0))}
```

The distribution of `nContact` looks like this:

```{r nContact2, echo=FALSE}
library(ggplot2)
library(dplyr)
set.seed(900)
data = data.frame(N=abs(round(rnorm(200, 0.5, 1), 0)))

data = data %>% group_by(N) %>% summarise(freq=length(N)/200)

ggplot(data=data, aes(x=as.factor(N), y=freq)) + geom_bar(stat="identity") + theme_minimal() + labs(x="nContact",y="Frequency")

```

At each time and for each infected host, `nContact` will be drawn anew. Remember that `nContact`, like the other core function has to be function of `t`, even if `t` is not used. Since `nContact` is constant here, there is no use of the "absolute" time of the simulation, the location of the host, nor individual-based parameters. So `param.nContact = NA`, `timeDep.nContact=FALSE` and `diff.nContact=FALSE`.

### pTrans

`pTrans` will be in the form of a threshold function: before a certain amount of time since initial infection, the host does not transmit (incubation time, we will call `t_incub`), and after that time, it will transmit with a certain (constant) probability (we will call `p_max`). This function will be dependant of the time since the host's infection `t`.

```{r pTrans1, eval = FALSE}
p_Trans_fct <- function(t, p_max, t_incub){
    if(t < t_incub){p=0}
    if(t >= t_incub){p=p_max}
    return(p)
}
```

Because each host is different (slightly different biotic and abiotic factors), you can expect each host to show differences in the dynamics of infection, and hence probability of transmission over time. Thus, `t_incub` and `p_max` will be sampled for each host individually according to a certain distribution. `t_incub` will be sampled from a normal distribution of $mean$ 7 and $sd$ 1, while `p_max` will be sampled from a beta distribution of parameters $\alpha$ 5 and $\beta$ 2.

```{r pTrans2, eval = FALSE}
t_incub_fct <- function(x){rnorm(x,mean = 7,sd=1)}
p_max_fct <- function(x){rbeta(x,shape1 = 5,shape2=2)}
```

Note that here `t_incub` and `p_max` are function of `x` and not `t` (they are not core function but individual-based parameters), and `x` enters the function as the number of draws to make.

Taken together, the profile for pTrans for a subset of 200 individuals in the population will look like this:

```{r pTrans3, echo=FALSE}
library(ggplot2)
library(dplyr)

set.seed(506)

p_Trans_fct <- function(t, p_max, t_incub){
    if(t < t_incub){p=0}
    if(t >= t_incub){p=p_max}
    return(p)
}

t_incub_fct <- function(x){rnorm(x,mean = 7,sd=1)}
p_max_fct <- function(x){rbeta(x,shape1 = 5,shape2=2)}

data = data.frame(t_incub=t_incub_fct(200),p_max=p_max_fct(200),host=paste0("H-",1:200))

t=c(0:12)
data3=NULL
for(t in 0:15){
  data2 = data %>% group_by(host) %>% mutate(proba=p_Trans_fct(t=t,p_max=p_max, t_incub=t_incub))
  data2$t = t
  data3 = rbind(data3, data2)
}

ggplot(data=data3, aes(x=t, y=proba,group=host)) + geom_line(color="grey60") + theme_minimal() + labs(x="Time since infection (t)",y="pTrans")
```

Since `pTrans` is not dependent of the "absolute" time of the simulation nor it is dependent of the hosts location. So `timeDep.pTrans=FALSE` and `diff.pTrans=FALSE`. However, since we have ever individual-based parameters, we have to provide a `param.pTrans` as a list of functions. The names of each element of the list should have the same name that the core function (pTrans here) uses as argument, such as here:

```{r pTrans4, eval = FALSE}
t_incub_fct <- function(x){rnorm(x,mean = 7,sd=1)}
p_max_fct <- function(x){rbeta(x,shape1 = 5,shape2=2)}

param_pTrans = list(p_max=p_max_fct,t_incub=t_incub_fct)
```

### Running

Once nosoiSim is set up, you can run the simulation (here the "seed" ensures that you will get the same results as in this tutorial).

```{r setupF}
library(nosoi)

#Transition matrix
transition.matrix <- matrix(c(0,0.2,0.4,0.5,0,0.6,0.5,0.8,0),nrow = 3, ncol = 3,dimnames=list(c("A","B","C"),c("A","B","C")))

#pExit
p_Exit_fct  <- function(t,current.in){
  if(current.in=="A"){return(0.02)}
  if(current.in=="B"){return(0.05)}
  if(current.in=="C"){return(0.1)}}

#pMove
p_Move_fct  <- function(t){return(0.1)}

#nContact
n_contact_fct = function(t){abs(round(rnorm(1, 0.5, 1), 0))}

#pTrans
proba <- function(t,p_max,t_incub){
  if(t <= t_incub){p=0}
  if(t >= t_incub){p=p_max}
  return(p)
}

t_incub_fct <- function(x){rnorm(x,mean = 5,sd=1)}
p_max_fct <- function(x){rbeta(x,shape1 = 5,shape2=2)}

param_pTrans = list(p_max=p_max_fct,t_incub=t_incub_fct)

# Starting the simulation ------------------------------------

set.seed(846)
SimulationA <- nosoiSim(type="single", popStructure="discrete",
                        length.sim=300, max.infected=300, init.individuals=1, init.structure="A", 
                        
                        structure.matrix=transition.matrix,
                        
                        pExit = p_Exit_fct,
                        param.pExit = NA,
                        timeDep.pExit=FALSE,
                        diff.pExit=TRUE,
                        
                        pMove = p_Move_fct,
                        param.pMove = NA,
                        timeDep.pMove=FALSE,
                        diff.pMove=FALSE,
                        
                        nContact=n_contact_fct,
                        param.nContact=NA,
                        timeDep.nContact=FALSE,
                        diff.nContact=FALSE,

                        pTrans = proba,
                        param.pTrans = list(p_max=p_max_fct,t_incub=t_incub_fct),
                        timeDep.pTrans=FALSE,
                        diff.pTrans=FALSE,
                      
                        prefix.host="H",
                        print.progress=FALSE,
                        print.step=10)
```
Once the simulation is finished, it reports the number of time units for which the simulation has run (36), and the maximum number of infected hosts (307).
Note that the simulation has stopped here before reaching `length.sim` as it has crossed the `max.infected` threshold set at 300.

## Dual host

Setting up a dual host simulation is similar to the single host version, but each parameters has to be provided for both hosts. Host A will use the same parameters as the only host above did. Host B will have sightly different parameters:

### pExit.B

For `pExit.B`, we will chose a value that depends on the "absolute" time of the simulation, for example cyclic climatic conditions (temperature). In that case, the function's argument should be `t` and `prestime` (the "absolute" time of the simulation), in that order:
```{r pExit1-dual, eval = FALSE}
p_Exit_fctB  <- function(t,prestime){(sin(prestime/(2*pi*10))+1)/16} #for a periodic function
```

The values of `pExit.B` across the "absolute time" of the simulation will be the following:

```{r pExit2-dual, echo = FALSE}
p_Exit_fctx <- function(x){(sin(x/(2*pi*10))+1)/16} #for a periodic function
ggplot(data=data.frame(x=0), aes(x=x)) + stat_function(fun=p_Exit_fctx) + theme_minimal() + labs(x="Absolute time (prestime)",y="pExit") + xlim(0,360)
```

Since `pExit.B` is dependent of the simulation's time, do not forget to turn `timeDep.pExit.B` to `TRUE`. Since there is no individual-based parameters nor influence of the hosts location, `param.pExit.B = NA` and `diff.pExit.B = NA`.

### pMove.B

We will assume here that hosts B do not move. `pMove.B` will then be set to `NA`.

```{r pMove.B, eval = FALSE}
p_Move_fct.B  <- NA
```

Since `pMove.B` is not dependent on the location, `diff.pMove.B=FALSE`. Similarly, there is no use of the "absolute" time of the simulation nor individual-based parameters, so `param.pMove.B = NA`, and `timeDep.pMove.B=FALSE`.

### nContact.B

For `nContact.B`, we will chose a constant function that will sample a value between a few that we chose, with a certain probability:

```{r nContact1.B, eval = FALSE}
n_contact_fct.B = function(t){sample(c(0,1,2),1,prob=c(0.6,0.3,0.1))}
```

The distribution of `nContact.B` looks like this:

```{r nContact2.B, echo=FALSE}
library(ggplot2)
library(dplyr)
set.seed(6059)
data = data.frame(N=sample(c(0,1,2),200,replace=TRUE,prob=c(0.6,0.3,0.1)))

data = data %>% group_by(N) %>% summarise(freq=length(N)/200)

ggplot(data=data, aes(x=as.factor(N), y=freq)) + geom_bar(stat="identity") + theme_minimal() + labs(x="nContact.B",y="Frequency")
```

At each time and for each infected host, `nContact.B` will be drawn anew. Remember that `nContact.B`, like the other core function has to be function of `t`, even if `t` is not used. Since `nContact.B` is constant here, there is no use of the "absolute" time of the simulation, the hosts location, nor individual-based parameters. So `param.nContact.B = NA`, `timeDep.nContact.B=FALSE` and `diff.nContact.B=FALSE`.

### pTrans.B

`pTrans.B` will be in the form of a gaussian function: it will its maximum value at a certain time point (mean) after initial infection and then will decrease until it reaches 0.

```{r pTrans1.B, eval = FALSE}
p_Trans_fct.B <- function(t, max.time){
  dnorm(t, mean=max.time, sd=2)*5
}
```

Because each host is here also different (slightly different biotic and abiotic factors), you can expect each host to show differences in the dynamics of infection, and hence probability of transmission over time. Thus, `max.time` will be sampled for each host individually according to a certain distribution. ``max.time` will be sampled from a normal distribution of parameters $mean$ 5 and $sd$ 1.

```{r pTrans2.B, eval = FALSE}
max.time_fct <- function(x){rnorm(x,mean = 5,sd=1)}
```

Note again that here `max.time` is function of `x` and not `t` (not core function but individual-based parameters), and `x` enters the function as the number of draws to make.

Taken together, the profile for `pTrans` for a subset of 200 individuals in the population will look like this:

```{r pTrans3.B, echo=FALSE}
library(ggplot2)
library(dplyr)
set.seed(6609)

p_Trans_fct <- function(t, max.time){
  dnorm(t, mean=max.time, sd=2)*5
}

max.time_fct <- function(x){rnorm(x,mean = 5,sd=1)}

data = data.frame(max.time=max.time_fct(200),host=paste0("H-",1:200))

t=c(0:12)
data3=NULL
for(t in 0:15){
  data2 = data %>% group_by(host) %>% mutate(proba=p_Trans_fct(t=t,max.time=max.time))
  data2$t = t
  data3 = rbind(data3, data2)
}

ggplot(data=data3, aes(x=t, y=proba,group=host)) + geom_line(color="grey60",alpha=0.3) + theme_minimal() + labs(x="Time since infection (t)",y="pTrans")
```

Since `pTrans.B` is not dependent of the "absolute" time of the simulation nor the hosts' location, `timeDep.pTrans.B=FALSE` and `diff.pTrans.B=FALSE`. However, since we have ever individual-based parameters, we have to provide a `param.pTrans` as a list of functions. The names of each element of the list should have the same name that the core function (`pTrans.B` here) uses as argument, such as here:

```{r pTrans4.B, eval = FALSE}
max.time_fct <- function(x){rnorm(x,mean = 5,sd=1)}

param_pTrans.B = list(max.time=max.time_fct)
```

### Running

Once nosoiSim is set up, you can run the simulation (here the "seed" ensures that you will get the same results as in this tutorial).

```{r setupF.B}
library(nosoi)

#Transition matrix
transition.matrix <- matrix(c(0,0.2,0.4,0.5,0,0.6,0.5,0.8,0),nrow = 3, ncol = 3,dimnames=list(c("A","B","C"),c("A","B","C")))

#Host A -----------------------------------

#pExit
p_Exit_fct  <- function(t,current.in){
  if(current.in=="A"){return(0.02)}
  if(current.in=="B"){return(0.05)}
  if(current.in=="C"){return(0.1)}}

#pMove
p_Move_fct  <- function(t){return(0.1)}

#nContact
n_contact_fct = function(t){abs(round(rnorm(1, 0.5, 1), 0))}

#pTrans
proba <- function(t,p_max,t_incub){
  if(t <= t_incub){p=0}
  if(t >= t_incub){p=p_max}
  return(p)
}

t_incub_fct <- function(x){rnorm(x,mean = 5,sd=1)}
p_max_fct <- function(x){rbeta(x,shape1 = 5,shape2=2)}

param_pTrans = list(p_max=p_max_fct,t_incub=t_incub_fct)

#Host B -----------------------------------

#pExit
p_Exit_fct.B  <- function(t,prestime){(sin(prestime/(2*pi*10))+1)/16}

#pMove
p_Move_fct.B  <- NA

#nContact
n_contact_fct.B = function(t){sample(c(0,1,2),1,prob=c(0.6,0.3,0.1))}

#pTrans
p_Trans_fct.B <- function(t, max.time){
  dnorm(t, mean=max.time, sd=2)*5
}
 
max.time_fct <- function(x){rnorm(x,mean = 5,sd=1)}

param_pTrans.B = list(max.time=max.time_fct)

# Starting the simulation ------------------------------------

set.seed(60)
SimulationA <- nosoiSim(type="dual", popStructure="discrete",
                        length.sim=300, 
                        max.infected.A=100,
                        max.infected.B=200,
                        init.individuals.A=1,
                        init.individuals.B=0,
                        init.structure.A="A",
                        init.structure.B=NA,
                        structure.matrix.A=transition.matrix,
                        structure.matrix.B=transition.matrix,
                        
                        pExit.A = p_Exit_fct,
                        param.pExit.A = NA,
                        timeDep.pExit.A=FALSE,
                        diff.pExit.A=TRUE,
                        
                        pMove.A = p_Move_fct,
                        param.pMove.A = NA,
                        timeDep.pMove.A=FALSE,
                        diff.pMove.A=FALSE,
                        
                        nContact.A=n_contact_fct,
                        param.nContact.A=NA,
                        timeDep.nContact.A=FALSE,
                        diff.nContact.A=FALSE,

                        pTrans.A = proba,
                        param.pTrans.A = list(p_max=p_max_fct,t_incub=t_incub_fct),
                        timeDep.pTrans.A=FALSE,
                        diff.pTrans.A=FALSE,
                        prefix.host.A="H",
                        
                        pExit.B = p_Exit_fct.B,
                        param.pExit.B = NA,
                        timeDep.pExit.B=TRUE,
                        diff.pExit.B=FALSE,
                        
                        pMove.B = p_Move_fct.B,
                        param.pMove.B = NA,
                        timeDep.pMove.B=FALSE,
                        diff.pMove.B=FALSE,
                        
                        nContact.B=n_contact_fct.B,
                        param.nContact.B=NA,
                        timeDep.nContact.B=FALSE,
                        diff.nContact.B=FALSE,

                        pTrans.B = p_Trans_fct.B,
                        param.pTrans.B = param_pTrans.B,
                        timeDep.pTrans.B=FALSE,
                        diff.pTrans.B=FALSE,
                        prefix.host.B="V",
                        
                        print.progress=FALSE)
```

Once the simulation is finished, it reports the number of time units for which the simulation has run (34), and the maximum number of infected hosts A (106) and hosts B (129).
Note that the simulation has stopped here before reaching `length.sim` as it has crossed the `max.infected.A` threshold set at 100.

# Going further

To analyse and visualise your `nosoi` simulation output, you can have a look on [this page](viz_stat.html).

A practical example using a dual host type of simulation with a discrete population structure is also available: 

- [Spread of dengue virus in a discrete space](dengue.html).
