---
title: "Spread of a pathogen in a homogeneous population"
author: "Sebastian Lequime"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{single-host-none}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`nosoi` can accommodate a wide range of epidemiological transmission scenarios, and doing so requires providing the proper parameter values.
This tutorial aims to illustrate how to set up a `nosoi` simulation for a "simple" case: a pathogen being transmitted among a single-host population without structure.

#Setting up the simulation

The wrapper function `nosoiSim` takes all the arguments that will be passed down to the core function, in the case of this tutorial: `singleNone` (for "single host, no structure").
We thus start by providing the options `type="single"` and `structure=FALSE` to set up the analysis:

```{r setupA, eval = FALSE}
SimulationA <- nosoiSim(type="single", structure=FALSE, ...)
```

This simulation (`singleNone`) type takes several arguments or options to be able to run, namely:

- `length.sim` 
- `max.infected` 
- `init.individuals` 
- `pExit` with `param.pExit` (optional)  and `timeDep.pExit` (optional)
- `timeContact` with `param.timeContact` (optional) and `timeDep.timeContact` (optional) 
- `pTrans` with `param.pTrans` and `timeDep.timeContact` (optional)
- `prefix.host` 

## General parameters

`length.sim`, `max.infected` and `init.individuals` are general parameters that define the simulation.
`length.sim` is the maximum number of time units (e.g. days, months, years, or another time unit of choice) during which the simulation will be run, whereas `max.infected` is the maximum number of individuals that can be infected during the simulation.
Both values need to be provided, and the simulation will thus run until either the time limit or the maximum number of infected individuals is reached.
`init.individuals` defines the number of individuals (starting at 1, should be an integer) that will start a transmission chain (you will have as many transmission chains as initial individuals that "seed" the epidemic process).

Here, we will run a simulation starting with 1 individual, for a maximum of 1.000 infected individuals and a maximum time of 300 days.

```{r setupB, eval = FALSE}
SimulationA <- nosoiSim(type="single", structure=FALSE,
                        length.sim=300, max.infected=1000, init.individuals=1, ...)
```

## Specific parameters

### `pExit`, `param.pExit` and `timeDep.pExit`

`pExit` is the first required fundamental parameter and provides a daily probability for a host to leave the simulation (either cured, died, etc.). 
To accommodate several scenarios, it can be a constant, time dependent or even individually parameterized, to include some stochasticity at the individual host level. 
In the latter case, you need to specify `param.pExit` as a list of functions.

In every case, the function provided, like all other similar function in `nosoi` has to be expressed as function of time `t`, even if time is not used to compute the probability.

Here are some possible functions for `pExit`:

```{r pExit, eval = FALSE}
p_Exit_fct  <- function(t){return(0.08)} #as a constant
p_Exit_fct  <- function(t){plogis(t,20,2)} #as a time-since infection dependent function

p_Exit_fct  <- function(t,pExit.param1){plogis(t,pExit.param1,2)} #as a time-since infection dependent function with an individually parameter
p_exit_param1 <- function(x){rnorm(x,mean = 10,sd=2)} #sampling one parameter for each infected individual
```

For a constant or time-dependent function, with no individual parameter, `param.pExit` is `NA`.
In the case of an individually parameterized `pExit`, the functions for the parameters have to be wrapped into a list, called `param.pExit`.
The name of each function in the list has to have the same name as the parameter it is sampling for `pExit`.
Here it will look like:

```{r param.pExit1, eval = FALSE}
param.pExit=NA
param.pExit=list(pExit.param1=p_exit_param1)
```

`timeDep.pExit` allows for `pExit` to be dependant of the "absolute time" of the simulation, to account, for example, for seasonality or other external time related covariates. 
By default, `timeDep.pExit` is set to `FALSE`. Here is a possible functions for `pExit` when `timeDep.pExit = TRUE`:

```{r timeDep.pExit, eval = FALSE}
p_Exit_fct  <- function(t,prestime){(sin(prestime/12)+1)/2} #for a periodic function
```

You can of course have an individually parameterized `pExit`, being both dependant of the time since infection (`t`) and "absolute" time (`prestime`), such as:

```{r timeDep.pExit2, eval = FALSE}
p_Exit_fct  <- function(t,prestime,pExit.param1){((sin(prestime/12)+1)/2)*plogis(t,pExit.param1,2)}
p_exit_param1 <- function(x){rnorm(x,mean = 10,sd=2)}
param.pExit=list(pExit.param1=p_exit_param1)
```

In any case, time since infection (`t`) and "absolute" time (`prestime`) should ALWAYS be designated by (`t`) and (`prestime`) respectively, and be in that order and first of the function parameters.

### `timeContact`, `param.timeContact` and `timeDep.timeContact`

`timeContact` represents the number (as an integer) of potentially infectious contacts an infected hosts can encounter per unit of time. At each time point, a number of contacts will be determined for each active host in the simulation.
The number of contacts (i.e. the output of your function) has to be an integer and can be set to zero. As this number might vary, it has to be expressed as a function of `t` (uniquely), but it can take various forms. Here are some examples:

```{r timeContact, eval = FALSE}
time_contact_fct = function(t){return(2)} #to get a constant number of contacts for each hosts, here 2.
time_contact_fct = function(t){sample(c(0,1,2,3),1,prob=c(0.4,0.3,0.2,0.1))} #to get a varying number of contacts for each hosts.
time_contact_fct = function(t){round(rnorm(1, 3, 1), 0)} #another example of varying number of contacts for each hosts.

time_contact_fct = function(t,prestime){
  if(prestime < 10){round(rnorm(1, 9, 1), 0)}
  if(prestime >= 10){round(rnorm(1, 3, 1), 0)}
} #for an absolute time dependent value
```

For a constant or time-dependent function, with no individual parameter, `param.timeContact` is `NA`.
In the case of an individually parameterized `timeContact`, the functions for the parameters have to be wrapped into a list, called `param.timeContact`.
The name of each function in the list has to have the same name as the parameter it is sampling for `timeContact`.
`timeDep.timeContact` allows for `timeContact` to be dependant of the "absolute time" of the simulation, to account, for example, for seasonality or other external time related covariates. 

### `pTrans`, `param.pTrans` and `timeDep.pTrans`

The `pTrans` function is the heart of the transmission process and represents the probability of transmission over time (when a contact occurs).
This function returns the probability (i.e. a value between 0 and 1) dependent on the time (`t`) since the host's infection.
It is intended to work with a set of parameters (at least 1) defined by `param.pTrans`, that will characterize the dynamics of the transmission probability over time, as well as introducing some individual based stochasticity.
Since each host is different (because of slightly different biotic and abiotic factors), you can expect each host to show differences in the dynamics of infection, and hence probability of transmission over time.
The parameters defined in `param.pTrans` are hence also functions that will sample parameter values (`param.pTrans` is a list of functions).
`timeDep.pTrans` allows for `pTrans` to be dependant of the "absolute time" of the simulation, to account, for example, for seasonality or other external time related covariates. 

As an example, let's consider the following function for `pTrans`, in the form of a threshold function:

```{r pTrans, eval = FALSE}
proba <- function(t,p_max,t_incub){
    if(t < t_incub){p=0}
    if(t >= t_incub){p=p_max}
    return(p)
}
```

This function returns a transmission probability `p` given the time since infection `t` and two parameters, `p_max`, the maximum transmission probability and `t_incub`, the length of the incubation period (the threshold).
If `t` is bigger than `t_incub`, then the transmission probability is `p_max`; if not, it is equal to 0.

`p_max` and `t_incub` need to be specified.
If we fix the values, for example to 0.8 and 5 units of time respectively, we assume that each host will react the same to the infection, as if they were perfect clones.
To more accurately match reality, `nosoi` forces you to sample the values of each parameters from a function, such as:

```{r param.pTrans1, eval = FALSE}
p_max_fct <- function(x){rbeta(x,shape1 = 5,shape2=2)}
t_incub_fct <- function(x){rnorm(x,mean = 5,sd=1)}
```

Here, `p_max` will be sampled from a beta distribution, with shape parameters 5 and 2, and `t_incub` will be sampled from a Gaussian distribution, with mean 5 and standard deviation 1.
`x` will be the number of draws to perform.
If you want to fix the value of one or all parameters, you can use a sampling function that will always give you the same value, such as:
```{r param.pTrans1bis, eval = FALSE}
t_incub_fct <- function(x){rep(5,x)} #constant parameter value.
```

In any case, the functions for `pTrans` parameters have to be wrapped into a list, called `param.pTrans`.
The name of each function in the list has to have the same name as the parameter it is sampling for `pTrans`.
Here, it will look like:

```{r param.pTrans2, eval = FALSE}
list(p_max=p_max_fct, t_incub=t_incub_fct)
```

At this step, the `nosoi` function should look like this:

```{r setupE, eval = FALSE}
SimulationA <- nosoiSim(type="single", structure=FALSE,
                        length.sim=300, max.infected=1000, init.individuals=1, 
                        timeContact=time_contact_fct,
                        param.timeContact=NA,
                        timeDep.timeContact=FALSE,
                        pExit = p_Exit_fct,
                        param.pExit = list(pExit.param1=p_exit_param1),
                        timeDep.pExit=FALSE,
                        pTrans = proba,
                        param.pTrans = list(p_max=p_max_fct,t_incub=t_incub_fct),
                        timeDep.pTrans=FALSE,
                        ...)
```

### Miscellaneous

`prefix.host` allows you to define the first character(s) for the hosts' unique ID.
It will be followed by a hyphen and a unique number.
By default, `prefix.host` is "H" for "Host".

`progress.bar` allows you to get some information printed on the screen about the simulation while running. It will print something every `print.step`. By default, `progress.bar` is activated, with a `print.step = 10` (that you can change), but you may want to deactivate it by providing `progress.bar == FALSE`.

## Running Nosoi

Once nosoiSim is set up, you can run the simulation (here the "seed" ensures that you will get the same results as in this tutorial).

```{r setupF}
library(nosoi)
#timeContact
time_contact_fct = function(t){abs(round(rnorm(1, 0.5, 1), 0))}

#pTrans
pTrans_fct <- function(t,p_max,t_incub){
  if(t <= t_incub){p=0}
  if(t >= t_incub){p=p_max}
  return(p)
}
 
t_incub_fct <- function(x){rnorm(x,mean = 7,sd=1)}
p_max_fct <- function(x){rbeta(x,shape1 = 5,shape2=2)}
  
#pExit
p_Exit_fct  <- function(t){return(0.08)}

set.seed(805)
SimulationA <- nosoiSim(type="single", structure=FALSE,
                        length.sim=100, max.infected=100, init.individuals=1, 
                        timeContact=time_contact_fct,
                        param.timeContact=NA,
                        timeDep.timeContact=FALSE,
                        pExit = p_Exit_fct,
                        param.pExit = NA,
                        timeDep.pExit=FALSE,
                        pTrans = pTrans_fct,
                        param.pTrans = list(p_max=p_max_fct,t_incub=t_incub_fct),
                        timeDep.pTrans=FALSE,
                        prefix.host="H",
                        progress.bar=FALSE)
```
The simulation can stop before 100% of the progress bar is reached if the maximum number of infected hosts has been reached.
Once the simulation is finished, it reports the number of time units for which the simulation has run (40), and the maximum number of infected hosts (111).
Note that the simulation has stopped here before reaching `length.sim` as it has crossed the `max.infected` threshold set at 100.

## Nosoi output

*Nosoi* gives as its output a list contaning a `data.table` (a more efficient version of a `data.frame`) called `table.hosts` with the following structure:

```{r output}
head(SimulationA$table.hosts)
```

Each individual host that was or is infected during the course of the simulation is a line in the data.table.
`hosts.ID` is the unique identifier of the host, while `inf.by` indicate which host infected it.
As you can see, host `H-1` has been infected by `NA`; `H-1` is the individual that started the transmission chain.
`inf.time` is the time (in simulation units of time) when the host has been infected, and `out.time` is the time (in simulation units of time) when it left the simulation.
Here in the top 6 lines, no host has left the simulation (it ran only for 13 units of time).
`active` indicates if the host is still active in the simulation.
`p_max` and `t_incub` are the individually sampled parameters for `pTrans`.

### Visualisation

The transmission chain can be visualized using some of the packages dealing with networks in R.
I typically use `igraph` with `ggnetwork`, but this is really up to you.
Keep in mind that the more nodes (i.e. infected hosts) you have, the more messy (or difficult) the visualisation will be.

```{r visualisation, message=FALSE}
library(ggplot2)
library(viridis)
library(igraph)
library(ggnetwork)

# To show the graph, we need to invert column 1 and 2, and discard the first line (i.e. the first host, 
# because it shows "NA" as infecting host). We then provide *Nosoi* output as elements for the vertices (nodes).
graph.simA <- graph.data.frame(SimulationA$table.hosts[-1,c(2,1)],directed=T,vertices = SimulationA$table.hosts)

graph.simA.network = ggnetwork(graph.simA, layout = "fruchtermanreingold", cell.jitter = 0.75) #using ggnetwork to provide the layout

#plotting the network
ggplot(graph.simA.network, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_edges(color = "grey70",arrow = arrow(length = unit(0.3, "lines"), type = "open")) +
  geom_nodes(aes(color=inf.time)) + scale_color_viridis(name="Time of infection",option = "plasma") + 
  theme_blank()
```

